Improving the portuguese hyphenation rules

### 2023-04-28 article/article.tex
adding information about lefthyphenmin and righthyphenmin

% xelatex --shell-escape compared.tex 
% pandoc -s article.tex --pdf-engine=xelatex --pdf-engine-opt=-shell-escape -o article.html
\documentclass{article}
\usepackage[english]{babel}
%\usepackage{testhyphens}
\usepackage{multicol}
\usepackage{verbatim}
\usepackage{hyperref}
\usepackage{amssymb}% http://ctan.org/pkg/amssymb
\usepackage{pifont}% http://ctan.org/pkg/pifont
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\usepackage{longtable}
\usepackage[style=alphabetic, backend=biber]{biblatex}
\addbibresource{references.bib}
\usepackage[inline]{enumitem}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{cleveref}

@@

\title{Improving the portuguese hyphenation rules}
\author{Leonardo Araújo \and Aline de Lima Benevides}
\begin{document}
\maketitle

@@

\section{Introduction}
Hyphenation in text wrapping was not used for a long time. Words should fit
entirely in a line, or they would be broken in arbitrary places. First no
marker at all was used to sign the wrapping of a word, what could create
confusion and undesirable meaning. Therefore, ortographers advocated for the
use of a sign to indicate this break. Portuguese faced the same gradual
introduction of a hyphenation sign to mark words wrapping across lines. Even
though the usage of a hyphenation sign ($=$) was advocated by orthographers
\cite{gandavo1574}, few documents used such sign until the end of the 18th
century \cite{araujo2015}.

@@

The effective hyphenation of words by \TeX{} will actually depends on the following factors:
\begin{enumerate*}[label=\arabic*)]
    \item document language, which will determine which set of patterns to apply;
    \item characters used, since some might block hyphenation at their edges;
    \item the value of the internal variables \verb|lefthyphenmin| and \verb|righthyphenmin|,
        which defines the minimum sequence length of characters at the left and right borders
        before any hyphenation is allowed.
\end{enumerate*}

@@

\section{Patterns for \TeX{} hyphenation} 
The patterns used in \TeX{} hyphenation are of the form: \texttt{4z1z2}. Each
pattern is made of a sequence of letters and numbers. Odd numbers indicate a
good hyphenation point, whereas even numbers indicate a bad place to break.
The given example states that the sequence has a good breaking point between
the first and the second \emph{z} and an hyphenation should be inhibited before
the first \emph{z} and after the second \emph{z}. For example, the hyphenation
of the word \emph{piz-za}, \emph{fiz-zle} and \emph{mezzanine} use this rule,
where we see the hyphen placed between the two \emph{z}'s and no hyphen before,
nor after the \emph{z}'s. The pattern may also use period symbol (\emph{.}) to
indicate word boundaries. The pattern \texttt{.s2h2} applies to beginning of
words, implying that the \emph{s} and the \emph{h} should stick together in
beginning of a word and an hyphenation should also be inhibited after the
\emph{h}. For example, this pattern is used in \emph{Sher-lock}.

@@

\section{PATGEN}
PATGEN uses a list of hyphenated words to define rules at different
levels and lengths, based on the patterns found in the data. It start with
short patterns and increases until a maximum pattern length, allowed by the
user, is reached. As it advances, creating longer patterns, it also establishes
exceptions. In some cases, the analysis of long patterns might be necessary,
since some hyphenation points might depend on characters far away from the
breaking point\footnote{Some examples of hyphenation dependency on characters
far from the break point: \emph{dem-o-crat} and \emph{de-moc-ra-cy};
\emph{as-pi-rin} and \emph{aspir-ing}; \emph{de-mon-stra-tive} and
\emph{dem-on-stra-tion}.}.

@@

@@

Systematizing the rules that guide syllable boundarie and, consequently,
hyphenation in Portuguese proved to be fundamental for the improvement of the
\TeX{} hyphenator rules, since they indicate the possible or prohibited contexts
of hyphenation. Expomos, a seguir, uma análise comparativa...

>>>>

% xelatex --shell-escape compared.tex 
% pandoc -s article.tex --pdf-engine=xelatex --pdf-engine-opt=-shell-escape -o article.html
\documentclass{article}
\usepackage[english]{babel}
%\usepackage{testhyphens}
\usepackage{multicol}
\usepackage{verbatim}
\usepackage{hyperref}
\usepackage{amssymb}% http://ctan.org/pkg/amssymb
\usepackage{pifont}% http://ctan.org/pkg/pifont
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\usepackage{longtable}
\usepackage[style=alphabetic, backend=biber]{biblatex}
\addbibresource{references.bib}
\usepackage[inline]{enumitem}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{fancyvrb}

@@

\title{Improving the portuguese hyphenation rules}
\author{Leonardo Araújo \and Aline de Lima Benevides}
\begin{document}
\VerbatimFootnotes
\maketitle

@@

\section{Introduction}\label{sec-intro}
Hyphenation in text wrapping was not used for a long time. Words should fit
entirely in a line, or they would be broken in arbitrary places. First no
marker at all was used to sign the wrapping of a word, what could create
confusion and undesirable meaning. Therefore, ortographers advocated for the
use of a sign to indicate this break. Portuguese faced the same gradual
introduction of a hyphenation sign to mark words wrapping across lines. Even
though the usage of a hyphenation sign ($=$) was advocated by orthographers
\cite{gandavo1574}, few documents used such sign until the end of the 18th
century \cite{araujo2015}.

@@

The effective hyphenation of words by \TeX{} will actually depends on the following factors:
\begin{enumerate*}[label=\arabic*)]
    \item document language, which will determine which set of patterns to apply;
    \item characters used, since some might block hyphenation at their edges;
    \item the value of the internal variables \verb|lefthyphenmin| and \verb|righthyphenmin|\footnote{
	The variables \verb|lefthyphenmin| and \verb|righthyphenmin| are language dependant and
	are defined in \emph{tlpobj} files (\verb|/usr/local/texlive/20XX/tlpkg/tlpobj/hyphen-xxxxxx.tlpobj|). 
	Default values varies in the range from 1 to 3. 
    	English and Portuguese, for example, use \verb|lefthyphenmin=2| and \verb|righthyphenmin=3|.},
	% ls /usr/local/texlive/2023/tlpkg/tlpobj/*.tlpobj | while read -r filename; do grep -Po "lefthyphenmin=[0-9]\s+righthyphenmin=[0-9]" $filename; done | tr '\t' ' ' | tr -s ' ' | sort | uniq -c
	%      17 lefthyphenmin=1 righthyphenmin=1
        %       6 lefthyphenmin=1 righthyphenmin=2
        %       1 lefthyphenmin=1 righthyphenmin=3
        %      49 lefthyphenmin=2 righthyphenmin=2
        %       9 lefthyphenmin=2 righthyphenmin=3
        which defines the minimum sequence length of characters at the left and right borders
        before any hyphenation is allowed.
\end{enumerate*}

@@

\section{Patterns for \TeX{} hyphenation}\label{sec-patt4TeX} 
The patterns used in \TeX{} hyphenation are of the form: \texttt{4z1z2}. Each
pattern is made of a sequence of letters and numbers. Odd numbers indicate a
good hyphenation point, whereas even numbers indicate a bad place to break.
The given example states that the sequence has a good breaking point between
the first and the second \emph{z} and an hyphenation should be inhibited before
the first \emph{z} and after the second \emph{z}. For example, the hyphenation
of the word \emph{piz-za}, \emph{fiz-zle} and \emph{mezzanine} use this rule,
where we see the hyphen placed between the two \emph{z}'s and no hyphen before,
nor after the \emph{z}'s. The pattern may also use period symbol (\emph{.}) to
indicate word boundaries. The pattern \texttt{.s2h2} applies to beginning of
words, implying that the \emph{s} and the \emph{h} should stick together in
beginning of a word and an hyphenation should also be inhibited after the
\emph{h}. For example, this pattern is used in \emph{Sher-lock}.

@@

\section{PATGEN}\label{sec-patgen}
PATGEN uses a list of hyphenated words to define rules at different
levels and lengths, based on the patterns found in the data. It start with
short patterns and increases until a maximum pattern length, allowed by the
user, is reached. As it advances, creating longer patterns, it also establishes
exceptions. In some cases, the analysis of long patterns might be necessary,
since some hyphenation points might depend on characters far away from the
breaking point\footnote{Some examples of hyphenation dependency on characters
far from the break point: \emph{dem-o-crat} and \emph{de-moc-ra-cy};
\emph{as-pi-rin} and \emph{aspir-ing}; \emph{de-mon-stra-tive} and
\emph{dem-on-stra-tion}.}.

@@

Note that rules \ref{rule-singlev} and \ref{rule-orphan} are fiddling rules in \TeX{} hyphenation, 
since \TeX{} already has directives to avoid hyphenated widows and orphans. 
As mentioned in \Cref{sec-intro}, the variables \verb|lefthyphenmin| and \verb|righthyphenmin|
are used to control the minimum length for fragments of hyphenated words.
This is the ground why we will not consider rules \ref{rule-singlev} and \ref{rule-orphan}
when analysing and elaborating the hyphenation rules.

@@

Systematizing the rules that guide syllable boundaries and, consequently,
hyphenation in Portuguese proved to be fundamental for the improvement of the
\TeX{} hyphenator rules, since they indicate the possible or prohibited contexts
of hyphenation. Expomos, a seguir, uma análise comparativa...
